---
title: "Adaptive Rejection Sampling"
author: "Dmitry Zotikov"
date: "March 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
set.seed(1)
```

# Toy example: моделирование $\mathrm{N}(0,1)$

```{r, eval=FALSE, include=FALSE}
f <- function(x) x^2
x <- 4
numericDeriv(quote(f(x)), c('x'))
```

```{r}
library('numDeriv')
f <- function(x) x^2
grad(f, 4)
```

TODO: use `cut`?

```{r}
p <- function(x) dnorm(x, -2, 1)
h <- function(x) log(p(x)) 
curve(h, -10, 10)
```

```{r}
k <- 4  # partition rank
tt.bounds <- c(-3, 3)
tt <- seq(tt.bounds[1], tt.bounds[2], length.out=k)  # partition itself
tt
hh <- h(tt)
hh.deriv <- grad(h, tt)
```

$\mathbf{z}=(z_{0},\ldots,z_{k})$ так что `zz` длины $k+1$ индексируется как `j. <- j+1`: если $z_{j},\ j\in0:k$, то $j_{\cdot}\in1:(k+1)$ и `zz[j.] == zz[j+1]`.

```{r}
zz <- numeric(k+1)
compute.z.j <- function(j, tt, hh, hh.deriv) {
  if (j == 0) {
    return(-Inf)
  } else if (j == length(tt)) {
    return(Inf)
  } else {
    (hh[j+1] - hh[j] - tt[j+1] * hh.deriv[j+1] + tt[j] * hh.deriv[j]) /
      (hh.deriv[j] - hh.deriv[j+1])
  }
}
for (j in 0:k) {
  j. <- j + 1
  zz[j.] <- compute.z.j(j, tt, hh, hh.deriv)
}
zz
```

```{r}
interval.idx <- function(x, xx, tol=10^-6) {
  if (x < min(xx) || x > max(xx)) {
    warning('Outside of partition boundaries')
    j <- NULL
  } else if (x == -Inf || abs(xx[1] - x) < tol) {
    j <- 1  # at the lower boundary -- a special case
  } else if (x == Inf) {
    j <- length(xx) - 1
  } else {
    j <- which(abs(xx - x) < tol) - 1 # at parition points
    if (length(j) == 0) {
      j <- which.max(xx > x) - 1
    }
  }
  return(j)
}
```

```{r, warning=FALSE}
for (x in c(-Inf, -4, -3, -2, -1, 0, 1, 2, 3, 4, Inf)) {
  cat(x, '->', interval.idx(x, tt), '\n')
}
for (x in c(-Inf, -3, -2, -1, 0, 1, 2, 3, Inf)) {
  cat(x, '->', interval.idx(x, zz), '\n')
}
```

```{r}
u <- function(x, tt, hh, hh.deriv, zz, tol=10^-6) {
  # cat('x =', x, '\n')
  if (x %in% c(-Inf, Inf)) {
    u <- -Inf
    # cat('u = -Inf\n')
  } else {
    j <- interval.idx(x, zz, tol)
    # cat('j =', j, '\n')
    u <- hh[j] + hh.deriv[j] * (x - tt[j])
  }
  return(u)
}
```

```{r}
for (x in c(-Inf, -3, -2, -1, 0, 1, 2, 3, Inf)) {
  print(u(x, tt, hh, hh.deriv, zz))
}
```

$\mathbf{u}=(u_{0},\ldots,u_{k})=(u(z_{0}),\ldots,u(z_{k})$ тоже индексируется `j.`

```{r}
uu <- sapply(zz, function(z) u(z, tt, hh, hh.deriv, zz))
uu
```

```{r}
compute.C.summands <- function(uu, hh.deriv) {
  k <- length(uu) - 1
  (exp(uu[2:(k+1)]) - exp(uu[1:k])) / hh.deriv
}
C.summands <- compute.C.summands(uu, hh.deriv)
C.summands
C <- sum(C.summands)
C
cdf.Q <- function(z, zz, C, tol=10^-6) {
  # cat('z =', z, '\n')
  if (z == -Inf) {
    return(0)
  }
  j.max <- interval.idx(z, zz, tol)
  # cat('jj =', 1:j.max, '\n')
  sum(C.summands[1:j.max]) / C
}
```

```{r}
for (z in c(-Inf, -4, -3, -2, -1, 0, 1, 2, 3, 4, Inf)) {
  cat('cdf.Q(', z, ') = ', cdf.Q(z, zz, C) ,'\n', sep='')
}
```

FIXME: NaN если $\alpha < F(z_1)$ потому что `zz[0] = -Inf`.

```{r}
sample.Q <- function(hh.deriv, zz, uu, C) {
  alpha <- runif(1)
  # cat('alpha =', alpha, '\n')
  FF <- sapply(zz, function(z) cdf.Q(z, zz, C))
  i. <- interval.idx(alpha, FF)
  # FIXME: HACK: what if alpha > FF?
  if (length(i.) == 0) {
    i. <- length(FF)
  }
  i <- i. - 1
  # cat('FF = ', FF, '\n')
  # cat('i =', i, '\n')
  zz[i.] + log(1 + hh.deriv[i+1] * C * (alpha - cdf.Q(zz[i.], zz, C)) / exp(uu[i.])) / hh.deriv[i+1]
}
sample.Q(hh.deriv, zz, uu, C)
```

```{r, eval=FALSE, include=FALSE}
xx <- replicate(1000, sample.Q(hh.deriv, zz, uu, C))
sum(is.nan(xx))
shapiro.test(xx)
hist(xx, freq=FALSE)
```

## Rejection sampling (upper hull)

```{r}
rs.maj <- function(tt, hh, hh.deriv, zz, uu, C) {
  i <- 1
  while (TRUE) {
    while (is.nan(eta <- sample.Q(hh.deriv, zz, uu, C))) {}
    # eta <- sample.Q(cdf.Q, zz, uu, C, hh.deriv)
    alpha <- runif(1)
    if (alpha <= exp(h(eta) - u(eta, tt, hh, hh.deriv, zz))) {
      return(c(eta, i))
    }
    i <- i+1
  }
}
```

```{r}
rs.runner <- function(n, rs) {
  etas <- numeric(n)
  n.iters <- numeric(n)
  
  low.n <- 0
  up.n <- 0
  for (i in 1:n) {
    res <- rs()
    etas[i] <- res[1]
    n.iters[i] <- res[2]
    if (length(res) > 2) {
      if (res[3] == -1) {
        low.n <- low.n + 1
      } else if (res[3] == 1) {
        up.n <- up.n + 1
      } else {
        warning(paste('Unknown convergence type:', res[3]))
      }
    }
  }
  cat('Avg. # of iterations:', sum(n.iters) / n, '\n')
  if (length(res) > 2) {
    cat('% lower hull:', low.n / n, '\n')
    cat('% upper hull:', up.n / n, '\n')
  }
  return(etas)
}
```

```{r}
set.seed(1)
xx <- rs.runner(1000, function() rs.maj(tt, hh, hh.deriv, zz, uu, C))
hist(xx, freq=FALSE)
shapiro.test(xx)
```

## Rejection sampling (upper + lower hulls)

```{r}
ell <- function(x, tt, hh) {
  if (x < tt[1] || x > tt[length(tt)]) {
    return(-Inf)
  }
  j <- interval.idx(x, tt)
  ((tt[j+1] - x) * hh[j] + (x - tt[j]) * hh[j+1]) / (tt[j+1] - tt[j])
}
```

```{r}
for (x in c(-Inf, -4:4, Inf)) {
  cat('ell(', x, ') = ', ell(x, tt, hh) ,'\n', sep='')
}
```

```{r}
rs <- function(tt, hh, hh.deriv, zz, uu, C) {
  i <- 1
  while (TRUE) {
    while (is.nan(eta <- sample.Q(hh.deriv, zz, uu, C))) {}
    alpha <- runif(1)
    u.eta <- u(eta, tt, hh, hh.deriv, zz)
    if (alpha <= exp(ell(eta, tt, hh) - u.eta)) {
      return(c(eta, i, -1))
    } else if (alpha <= exp(h(eta) - u.eta)) {
      return(c(eta, i, 1))
    }
    i <- i+1
  }
}
```

```{r}
set.seed(1)
xx <- rs.runner(1000, function() rs(tt, hh, hh.deriv, zz, uu, C))
hist(xx, freq=FALSE)
shapiro.test(xx)
```

## Adaptive Rejection Sampling

```{r}
ars <- function(tt, hh, hh.deriv, zz, uu, C) {
  i <- 1
  while (TRUE) {
    while (is.nan(eta <- sample.Q(hh.deriv, zz, uu, C))) {}
    # eta <- sample.Q(hh.deriv, zz, uu, C)
    alpha <- runif(1)
    u.eta <- u(eta, tt, hh, hh.deriv, zz)
    if (alpha <= exp(ell(eta, tt, hh) - u.eta)) {
      return(c(eta, i, -1))
    } else {
      h.eta <- h(eta)
      if (eta > tt.bounds[1] && eta < tt.bounds[2]) {
        # FIXME: really?
        tt <- sort(c(tt, eta))
        # if (length(tt) == 27) browser()
        j <- interval.idx(eta, tt) + 1
        j. <- j+1
        hh <- append(hh, h.eta, after=j-1)
        hh.deriv <- append(hh.deriv, grad(h, eta), after=j-1)
        z.j <- compute.z.j(j, tt, hh, hh.deriv)
        zz <- append(zz, z.j, after=j.-1)
        uu <- append(uu, u(z.j, tt, hh, hh.deriv, zz), after=j.-1)
        C.summands <- compute.C.summands(uu, hh.deriv)
        C <- sum(C.summands)
        tt <<- tt
        hh <<- hh
        hh.deriv <<- hh.deriv
        zz <<- zz
        uu <<- uu
        C.summands <<- C.summands
        C <<- C
      }
      if (alpha <= exp(h.eta - u.eta)) {
        return(c(eta, i, 1))
      }
    }
    i <- i+1
  }
}
```

```{r, eval=FALSE, include=FALSE}
# ars(tt, hh, hh.deriv, zz, uu, C)
```

```{r}
set.seed(1)
xx <- rs.runner(1000, function() ars(tt, hh, hh.deriv, zz, uu, C))
hist(xx, freq=FALSE)
shapiro.test(xx)
```

```{r, fig.height=8, fig.width=10}
par(mfrow=c(2,3))
pars <- list(tt, hh, hh.deriv, zz, uu)
p.names <- c('x.j', 'h(x.j)', "h'(x.j)", 'z.j', 'u(z.j)')
for (i in 1:length(pars)) {
  plot(pars[[i]], type='b', main=p.names[i])
}
par(mfrow=c(1,1))
```

# Задание

Плотность распределения Бирнбаума--Саундерса $$p_{\mathrm{BS}}(x)=\frac{\sqrt{x}+\sqrt{1/x}}{2\gamma x} \phi\left(\frac{\sqrt{x}-\sqrt{1/x}}{\gamma}\right), \quad x>0,\gamma>0,\phi(x)=\mathrm{pdf}_{\mathrm{N}(0,1)}(x).$$

```{r}
p.bs <- function(x, gamma) {
  (sqrt(x) + sqrt(1/x)) / (2*gamma*x) * dnorm((sqrt(x) - sqrt(1/x)) / gamma)
}
```

```{r}
plot.f.par <- function(f, pp, from=0, to=5) {
  par(mfrow=c(2,2))
  for (p in pp) {
    f.pp <- function(x) f(x, p)
    curve(f.pp, from=from, to=to, main=p)
  }
  par(mfrow=c(1,1))
}

```

```{r, fig.height=8, fig.width=8}
plot.f.par(p.bs, c(0.1, 0.5, 1, 2))
```

```{r}
p.bs.log <- function(x, gamma) {
  log(p.bs(x, gamma))
}
plot.f.par(p.bs.log, c(0.1, 0.5, 1, 2))
```


$$\frac{\sqrt{x}+\sqrt{1/x}}{2\gamma x}$$
```{r, fig.height=8, fig.width=8}
f <- function(x, gamma) {
  (sqrt(x) + sqrt(1/x)) / (2*gamma*x)
}
plot.f.par(f, c(0.1, 0.5, 1, 2), to=1)
```

$$\phi\left(\frac{\sqrt{x}-\sqrt{1/x}}{\gamma}\right)$$
```{r, fig.height=8, fig.width=8}
g <- function(x, gamma) {
  dnorm((sqrt(x) - sqrt(1/x)) / gamma)
}
plot.f.par(g, c(0.1, 0.5, 1, 2), to=10)
```

$$\exp\left\{ -\frac{1}{2}\frac{x^{2}}{\gamma^{2}}\right\}$$

```{r, fig.height=8, fig.width=8}
g.1 <- function(x, gamma) {
  exp(-1/2 * x^2 / gamma^2)
}
plot.f.par(g.1, c(0.1, 0.5, 1, 2), to=10)
```

$$\exp\left\{ \frac{1}{\gamma^{2}}\right\}$$

```{r, fig.height=8, fig.width=8}
g.2 <- function(x, gamma) {
  exp(1 / gamma^2) * x/x
}
plot.f.par(g.2, c(0.1, 0.5, 1, 2), to=10)
```

$$\exp\left\{ -\frac{1}{2}\frac{1}{x\gamma^{2}}\right\}$$

```{r, fig.height=8, fig.width=8}
g.3 <- function(x, gamma) {
  exp(-1/2 * 1/(x * gamma^2))
}
plot.f.par(g.3, c(0.1, 0.5, 1, 2), to=10)
```